Type theories as quotient inductive-recursive types

### 2025-02-06 tex/main.tex
Polish S4.1

\begin{abstract}
  Type theory can be defined in a type theory as a quotient inductive-inductive type, but well-typed terms are littered with explicit coercions, i.e.\ \transp's along equality constructors, and proofs need to account for coercions painstakingly alongside the interesting part.
  This mess has been dubbed the `transport hell' but typically suppressed in presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim for shortening the gap using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
  As a case study, we investigate (parallel) substitution calculus and type theory with a universe and $\Pi$-types as quotient inductive-inductive-recursive types and compare ours with quotient inductive-inductive definitions.
\end{abstract}

@@

In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types.
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
%
This is helpful to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
%
As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with quotient inductive-inductive types alone).
\LT{terminology: strict vs.\ definitional vs.\ judgemental?}
%

@@

%
We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
Nevertheless, Agda can be convinced to accept our formal implementation, using some experimental flags.
Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.

@@

  For other two equality constructors, the arguments follow similarly.
\end{proof}
The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
\begin{proposition}[Coherence, II]
  For the following equality constructors $p : \sigma = \tau$
  \begin{alignat*}{5}
    \pi_1\beta    & : \pi_1(\sigma, t)  && = \sigma \qquad\qquad & \mathsf{concat} & : \sigma; (\tau, t) && = (\sigma;\tau), [\sigma]\, t \\
    \emptyctx\eta & : \sigma            && = \emptyctx & \pi\eta & : \sigma && = \pi_1\sigma, \pi_2\sigma
  \end{alignat*}
   the identities
  \danger
  \begin{align*}
  [\sigma]_{\Tel}\,\Xi & = [\tau]_{\Tel}\,\Xi 
                         & [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A \\
  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
%    &
%  [\sigma \upuparrows \Xi ]_{\Tm}\,t & =^{\lam{\gamma}\Tm\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)\;([\gamma \upuparrows \Xi]_{\Ty} A)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Tm}\,t
  \end{align*}
  hold for any telescope $\Xi : \Tel\,\Delta$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
\end{proposition}
\begin{proof}
  We demonstrate the proof for $\pi_1\beta: \pi_1(\sigma, t) = t$ and other three cases follow similarly.

  First, we prove the coherence for $[\blank]_{\Ty}\,A$ by induction on $A$, assuming the coherence for the other two identities $[\blank]_{\Tel}\, \Xi$ and $\blank \upuparrows \Xi$, which are then proved by induction on $\Xi$.
  \begin{enumerate}
    \item By induction on $A$, consider the following cases for $[\sigma\upuparrows\Xi]_{\Ty}\,A = [\tau \upuparrows \xi]_{\Ty}\,A$.
      \begin{description}
        \item[$A$ is $\UU$:] By definition
          \[
            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \quad\text{and}\quad [\sigma\upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}
          \]
          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by the congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
          Then, $\UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \simeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}$ holds by congruence for $\UU$.
          \LT[noinline]{I think the congruence for $\UU$ is true but not sure how precisely.}

@@

\subsection{... and other type formers} \label{subsec:SC+U+Pi+more}
\LT{%
\begin{enumerate}
  \item Discuss the unit type.
  \item Discuss the Boolean type.
  \item Discuss the type of natural numbers.
  \item Discuss the extensional identity type.
\end{enumerate}
}

\section{The eliminator, the standard model, and the equivalence to the QII definition of type theory}
\LT{%
\begin{enumerate}
  \item Introduce motives and methods for the definition in \Cref{subsec:SC+U+Pi}.
  \item The rule $\mathsf{Elim}_{\Ty}\,([ \sigma ]_{\Ty}\,A) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Ty}(\mathsf{Elim}_{\Ty}\,A)$ is not defined but provable. 
  \item $\mathsf{Elim}_{\Tm}\,([ \sigma ]_{\Tm}\,t) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Tm}(\mathsf{Elim}_{\Tm}\,t)$ is proved.
\end{enumerate}
}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}
After giving the QIIR definition of a type theory, we have to define how to \emph{eliminate} types $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ altogether to another types (called \emph{motives}) by functions (called \emph{methods}) subject to equations corresponding to point constructors and equality constructors (\cref{subsec:eliminator}).
Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart (\cref{subsec:equivalence}).

\subsection{Eliminator} \label{subsec:eliminator}
As the same as the QII definition~\cite{Altenkirch2016a}, the motives for the QIIR definition are
\begin{align*}
  \Ctx^A & : \Ctx \to \Set &
  \Sub^A & : \implicit{\Gamma, \Delta}\,\Ctx^A\;\Gamma \to \Ctx^A\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
  \Ty^A  & : \implicit{\Gamma}\,\Ctx^A\;\Gamma \to \Ty\;\Gamma \to \Set & 
  \Tm^A  & : \implicit{\Gamma, A}\; (\Gamma^A : \Ctx^A\;\Gamma) \to \Ty^A\;\Gamma^A\;A \to \Tm\;\Gamma\;A \to \Set
\end{align*}
and each method corresponds to each constructor, e.g., 
\[
  \emptyctx^A : \Ctx^A\;\emptyctx 
  \quad
  \blank ,^A \blank : \implicit{\Gamma, A}\;(\Gamma^A : \Ctx^A\;\Gamma) \to \Ty^A\;\Gamma^A\; A \to \Ctx^A\;(\Gamma, A)
\]
\subsection{The standard model} \label{subsec:std-model}
\subsection{The equivalence} \label{subsec:equivalence}
\section{Formal implementation}
\LT{Mention the NbE for SC here}
\LT{Compare some formal proofs with pen-and-paper proofs here}
\section{Discussion}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
\cite{Altenkirch2017}
\cite{Munoz1998}
\cite{Dybjer1996,Castellan2021}
\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}

>>>>

\begin{abstract}
  \LT[noinline]{Try to interest Thorsten Altenkirch in the abstract}
  Defining type theory in type theory is a long-standing challenge and a benchmark for the expressiveness of a proof assistant.
  Type theory consisting of well-typed terms only is defined using quotient inductive-inductive type by Altenkirch and Kaposi, but well-typed terms are littered with explicit coercions, i.e.\ \transp's along equality constructors, and proofs need to account for coercions painstakingly alongside the interesting part.
  This mess has been dubbed the `transport hell' but typically suppressed in presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim for shortening the gap using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
  As a case study, we investigate (parallel) substitution calculus and type theory with a universe and $\Pi$-types and compare ours with quotient inductive-inductive definitions.
\end{abstract}

@@

In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types.
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Dybjer~\cite{Dybjer1996} and Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
%
This is helpful to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
%
As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with quotient inductive-inductive types alone).
\LT{terminology: strict vs.\ definitional vs.\ judgemental?}
%

@@

%
We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
Nevertheless, Agda can be convinced to accept our formal implementation, using some experimental flags to introduce additional rewrite rules.\footnote{%
  As shown in \cite{Cockx2021}, adding non-terminating rewrite rules does not break consistency as long as these rules are consistent, so consistency is not an issue in our development.
}
Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.

@@

  For other two equality constructors, the arguments follow similarly.
\end{proof}
The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
\begin{proposition}[Coherence, II]
  \LT[noinline]{Josh Ko suggests us to follow the proof structure in our formal implementation.}
  For the following equality constructors $p : \sigma = \tau$
  \begin{alignat*}{5}
    \pi_1\beta    & : \pi_1(\sigma, t)  && = \sigma \qquad\qquad & \mathsf{concat} & : \sigma; (\tau, t) && = (\sigma;\tau), [\sigma]\, t \\
    \emptyctx\eta & : \sigma            && = \emptyctx & \pi\eta & : \sigma && = \pi_1\sigma, \pi_2\sigma
  \end{alignat*}
   the identities
  \danger
  \begin{align*}
  [\sigma]_{\Tel}\,\Xi & = [\tau]_{\Tel}\,\Xi 
                         & [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A \\
  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
%    &
%  [\sigma \upuparrows \Xi ]_{\Tm}\,t & =^{\lam{\gamma}\Tm\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)\;([\gamma \upuparrows \Xi]_{\Ty} A)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Tm}\,t
  \end{align*}
  hold for any telescope $\Xi : \Tel\,\Delta$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
\end{proposition}
\begin{proof}
  We show the statement for $\pi_1\beta: \pi_1(\sigma, t) = t$ and the other three cases follow similarly.

  First, we prove the coherence for $[\blank]_{\Ty}\,A$ by induction on $A$ and the coherence for the other two identities $[\blank]_{\Tel}\, \Xi$ and $\blank \upuparrows \Xi$ by induction on $\Xi$.
  \begin{enumerate}
    \item Consider the following cases for $[\sigma\upuparrows\Xi]_{\Ty}\,A = [\tau \upuparrows \xi]_{\Ty}\,A$ by induction on $A$.
      \begin{description}
        \item[$A$ is $\UU$:] By definition
          \[
            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \quad\text{and}\quad [\sigma\upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}
          \]
          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
          Then, $\UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \simeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}$ holds by congruence for $\UU$.
          \LT{Justify the use of congruence rule for QIIT, e.g. $\transfib{\Ty}{p}{\UU_{\Gamma}} = \UU_{\Delta}$ for $p : \Gamma = \Delta$}

@@

\subsection{... and other type formers} \label{subsec:SC+U+Pi+more}
\LT{%
\begin{enumerate}
  \item Discuss the extensional identity type.
  \item Discuss Coquand universes.
  \item Discuss the unit type.
  \item Discuss the Boolean type.
  \item Discuss the type of natural numbers.
\end{enumerate}
}

\section{The elimination principle and models}
\LT{Terminology: the eliminator principle vs.\ eliminator vs.\ induction principle?}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}
With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}) .
Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).

\subsection{The elimination principle} \label{subsec:eliminator}
To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify types
\begin{align*}
  \Ctx^M & : \Ctx \to \Set &
  \Sub^M & : \implicit{\Gamma, \Delta}\,\Ctx^M\;\Gamma \to \Ctx^M\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
  \Ty^M  & : \implicit{\Gamma}\,\Ctx^M\;\Gamma \to \Ty\;\Gamma \to \Set & 
  \Tm^M  & : \implicit{\Gamma, A}\; (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Tm\;\Gamma\;A \to \Set
\end{align*}
called the \emph{motivates} as the target types for the elimination principle:
\begin{alignat*}{5}
  \elim_{\Ctx} & : (\Gamma : \Ctx)   && \to \Ctx^M\;\Gamma \\
  \elim_{\Ty}  & : (A : \Ty\;\Gamma) && \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \\
  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\
  \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t.
\end{alignat*}

As inhabitants of these inductive types are formed by constructors, each constructor $c$ is interpreted by a corresponding functions $c^M$ indexed by $c$ such as
\begin{alignat*}{3}
  \emptyctx_{\Ctx}^M     & :                      && \Ctx^M\;\emptyctx \\
  \blank,_{\Ctx}^M\blank & : \implicit{\dots}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
  \emptysub_{\Sub}^M     & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
  \blank,_{\Sub}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
  \idS^M          & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
  \blank;\blank   & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Sub^M\;\Delta^M\;\Theta^M\;\tau \to \Sub^M\;\Gamma^M\;\Theta^M\;(\sigma;\tau)\\
  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t)
\end{alignat*}
and simultaneously defined functions are also given by corresponding functions 
\begin{alignat*}{3}
  [\blank]_{\Ty}^M\blank & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t)
\end{alignat*}
subject to the corresponding equations \emph{and function clauses}, e.g., 
\begin{alignat*}{5}
  [\idS]_{\Ty}^M & : \implicit{\dots} && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
  [;]_{\Ty}^M & : \implicit{\dots} && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
  [\idS]_{\Tm}^M & : \implicit{\dots} && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M
\end{alignat*}
called \emph{methods}.
For brevity, we omit the rest of methods and refer to the Agda implementation for their complete definition.
\LT[noinline]{Add a link.}

\begin{definition}
  An \emph{eliminator} for the QIIR definition of type theory consists of motives and methods above.
\end{definition}

\begin{remark}
To state these methods, we cannot avoid using transports in identities that would need transports in its QII definition of a type theory (e.g.\ $[\idS]_{\Tm}^M$) since identities (e.g.\ $[\idS]_{\Ty}^M$) are propositional just like equality constructors in its QII definition.
However, in the elimination principle for the QII definition, the situation is even worse: the above identity $[\idS]_{\Ty}^M$ \emph{already} requires $\alert{\transfib{\Ty^M\;\Gamma^M}{[\idS]_{\Ty}}{\color{black}[\idS^M]_{\Ty}^M\,A^M}}$ to coerce its type $\Ty^M\;\Gamma^M\;([\idS]\,A)$ to $\Ty^M\;\Gamma^M\;A$ to match the index $A$ of $A^M$, which are not needed in our QIIR definition.
That is, the eliminator for the QIIR definition requires fewer transports than its QII counterpart.
Nevertheless, the difference is insignificant, as both definitions would be derived from the signature of a quotient inductive-inductive(-recursive) type.
\end{remark}

Given an eliminator, the computation is given to the following function clauses
\begin{alignat*}{5}
  \elim_{\Ctx}\,\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
  \elim_{\Ctx}\,(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) \\
  \elim_{\Ty}\,\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
  \elim_{\Sub}\,\emptysub   & \reduce \emptysub^M_{\Sub} &&: \Sub^M\;\Gamma^M\;\emptyctx^M_{\Ctx}\;\emptyctx \\
                            & \vdots
\end{alignat*}
by interpreting each constructor using the corresponding method inductively, similar to its QII counterpart.
Here, the difference is that type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
In contrast, substitution operations $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}$ are functions in our QIIR definition that cannot be eliminated inductively, but they become propositional identities:
\begin{proposition}\label{prop:provable-function-clauses}
  Given the motives and the methods, we have
\begin{alignat*}{5}
  \elim_{\Ty}[] & : \implicit{\dots} & \elim_{\Ty}\,([\sigma]_{\Ty}\,A) & = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
  \elim_{\Tm}[] & : \implicit{\dots} & \elim_{\Tm}\,([\sigma]_{\Tm}\,t) & =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t)
\end{alignat*}
\end{proposition}
\LT{Add some remark regarding the proof?}

These identities are required to interpret constructors using definitional identities.
E.g., consider eliminating an extension $(\sigma, t)$ for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$, i.e.\ 
\[
  \elim_{\Sub}(\sigma, t) = (\elim_{\Sub}\,\sigma) ,^M_{\Sub} \; \alert{t^M}.
\]
In the QII definition, $t^M$ is just $\elim_{\Tm}\,t$ of type
\begin{equation} \label{eq:elim-subext}
  \Tm^M\;\Gamma^M\;(\elim_{\Ty}\;([\sigma]_{\Ty}\,A))\;t\judgeq \Tm^M\;\Gamma^M\;[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)\;t
\end{equation}
as the RHS of \eqref{eq:elim-subext} is indeed the type for the second argument of $\blank,^M_{\Sub}\blank$.
In the QIIR definition, on the contrary, the type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$ which is not definitionally equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ is in fact
\[
  (\elim_{\Sub}\,\sigma) ,_{\Sub}^M \alert{\transfib{\Tm^M\;(\elim_{\Ctx}\,\Gamma)}{\elim_{\Ty}[]}{\color{black}(\elim_{\Tm}\,t)}}
\]
requiring a transport of $\elim_{\Tm}\,t$ along the identity $\elim_{\Ty}[]$.

\begin{remark}
Note again that our inductive-recursive definition is not an instance of the general schema of inductive-recursion in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003} where simultaneously defined functions do not require corresponding interpretation as they target meta-theoretic types instead of inductive types being defined.
\end{remark}

\subsection{The standard model} \label{subsec:std-model}
Our QIIR definition of type theory has a standard model by giving the following motives and methods.

\subsection{The equivalence} \label{subsec:equivalence}
\section{Formal implementation}
\LT{Mention the NbE for SC here}
\LT{Compare some formal proofs with pen-and-paper proofs here}
\section{Discussion}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
\cite{Altenkirch2017}
\cite{Munoz1998}
\cite{Dybjer1996,Castellan2021}
\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}
